Conformal Counterfactual Explanations

### 2023-03-22 notebooks/proposal.qmd
some finetuning of intro examples

```{julia}
# Data:
counterfactual_data = load_mnist()
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
input_dim, n_obs = size(counterfactual_data.X)
M = load_mnist_mlp()
# Target:
factual_label = 8
x = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 3
factual = predict_label(M, counterfactual_data, x)[1]
# Search:
n_ce = 3
generator = GenericGenerator()
ces = generate_counterfactual(x, target, counterfactual_data, M, generator; num_counterfactuals=n_ce)
```

```{julia}
image_size = 200
p1 = plot(
    convert2image(MNIST, reshape(x,28,28)),
    axis=nothing, 
    size=(image_size, image_size),
    title="Factual"
)
plts = [p1]

counterfactuals = CounterfactualExplanations.counterfactual(ces)
phat = target_probs(ces)
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3))
    ce, _phat = (x[1],x[2])
    _title = "p(y=$(target)|x′)=$(round(_phat[1]; digits=3))"
    plt = plot(
        convert2image(MNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(image_size, image_size),
        title=_title
    )
    plts = [plts..., plt]
end
plt = plot(plts...; size=(image_size * (n_ce + 1),image_size), layout=(1,(n_ce + 1)))
savefig(plt, joinpath(www_path, "you_may_not_like_it.png"))
```

@@

>>>>

```{julia}
# Data:
counterfactual_data = load_mnist()
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
input_dim, n_obs = size(counterfactual_data.X)
M = load_mnist_mlp()

# Target:
factual_label = 8
x = reshape(X[:,rand(findall(predict_label(M, counterfactual_data).==factual_label))],input_dim,1)
target = 3
factual = predict_label(M, counterfactual_data, x)[1]
γ = 0.9
T = 50
```

```{julia}
# Search:
generator = GenericGenerator()
ce_wachter = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
generator = GreedyGenerator(η=5.0)
ce_jsma = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

```{julia}
p1 = plot(
    convert2image(MNIST, reshape(x,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

ces = zip([ce_wachter,ce_jsma])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3))
    ce, _phat = (x[1],x[2])
    _title = "p(y=$(target)|x′)=$(round(_phat[1]; digits=3))"
    plt = plot(
        convert2image(MNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
savefig(plt, joinpath(www_path, "you_may_not_like_it.png"))
```

@@

```{julia}
using CounterfactualExplanations.Models: load_mnist_vae
vae = load_mnist_vae()
vae_weak = load_mnist_vae(;strong=false)
Serialization.serialize(joinpath(output_path,"mnist_classifier.jls"), M)
Serialization.serialize(joinpath(output_path,"mnist_vae.jls"), vae)
Serialization.serialize(joinpath(output_path,"mnist_vae_weak.jls"), vae_weak)
```

```{julia}
# Define generator:
generator = REVISEGenerator(
  opt = Descent(0.1),
  λ=0.01
)
# Generate recourse:
counterfactual_data.generative_model = vae # assign generative model
ce = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
counterfactual_data = deepcopy(counterfactual_data)
counterfactual_data.generative_model = vae_weak
ce_weak = generate_counterfactual(
    x, target, counterfactual_data, M, generator;
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

```{julia}
ces = zip([ce,ce_weak])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
plts = [p1]
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3))
    ce, _phat = (x[1],x[2])
    _title = "p(y=$(target)|x′)=$(round(_phat[1]; digits=3))"
    plt = plot(
        convert2image(MNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
savefig(plt, joinpath(www_path, "surrogate_gone_wrong.png"))
```

### 2023-03-22 paper/paper.tex
some finetuning of intro examples

>>>>

\begin{figure}
  \centering
  \begin{minipage}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../www/you_may_not_like_it.png}
    \caption{You may not like it, but this is what stripped-down counterfactuals look like. Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using \citet{wachter2017counterfactual} (center); and a counterfactual produced using JSMA-based approach introduced by \citep{schut2021generating}.}\label{fig:adv}
  \end{minipage}\hfill
  \begin{minipage}[t]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../www/surrogate_gone_wrong.png}
    \caption{Using surrogates can improve plausibility, but also increases vulnerability. Counterfactuals for turning an 8 (eight) into a 3 (three): original image (left); counterfactual produced using REVISE \citep{joshi2019realistic} with a well-specified surrogate (center); and a counterfactual produced using REVISE \citep{joshi2019realistic} with a poorly specified surrogate (right).}\label{fig:vae}
  \end{minipage}
\end{figure}

\bibliography{../bib}
